---
title: "Intermediate R programming exercise solutions"
author: "Kim Dill-McFarland"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
mainfont: Arial
fontsize: 11 pt
editor_options:
  chunk_output_type: console
urlcolor: blue
---
# Setup
We will be working with the same data and packages as in the notes and main.R files.

```{r message=FALSE}
library(tidyverse)
library(lmerTest)
library(devtools)
library(roxygen2)
```

```{r message=FALSE}
raw_data <- read_csv(file="data/Saanich_Data.csv",
                     col_names=TRUE,
                     na=c("", "NA", "NAN", "ND"))

dat <- 
  raw_data %>%
  select(Cruise, Date, Depth, 
         Temperature, Salinity, Density, 
         WS_O2, WS_NO3, WS_H2S) %>%
  filter(Date >= "2008-02-01") %>%
  rename(O2=WS_O2, NO3=WS_NO3, H2S=WS_H2S) %>%
  mutate(Depth=Depth*1000)
```

# S3 objects
## Vectors
1. Assign x the value `"a"`. What are its class and mode?

```{r}
x <- "a"
class(x)
mode(x)
```

2. Give it dimensions `c(1,1)`. What are its class and mode?
```{r}
dim(x) <- c(1,1)
class(x)
mode(x)
```

## Data objects
1. Calculate a summary table of `dat`. What are its class and attributes?
```{r}
sum <- summary(dat)
class(sum)
attributes(sum)
```

2. Read in the raw data table `Saanich_Data.csv` using the base R function `read.table`. What are this object's class and attributes? Are they any different from the object create when we used `read_csv` to read in the same data?
```{r}
dat2 <- read.table("data/Saanich_Data.csv", sep=",")
class(dat2)
attributes(dat2)
```

## R list object
1. Calculate the `summary()` of `m1` and save it as `m2`.
```{r}
m1 <- lm(O2 ~ Depth, data=dat)
m2 <- summary(m1)
m2
```

2. What is the class and mode of `m2`?
```{r}
class(m2)
mode(m2)
```

3. Using a single line of code, pull out just the p-values from `m2`.
    - *Hint*: You will need to use both `$` and `[ ]`.
```{r}
m2$coefficients[,"Pr(>|t|)"]
```

# S4 objects
1. Compute and store the variance-covariance matrix of `m3` using `vcov()`.
```{r}
m3 <- lmer(O2 ~ Cruise + (0 + Cruise | Depth), dat)
vcov.m3 <- vcov(m3)
```

2. What class and mode is it?
```{r}
class(vcov.m3)
mode(vcov.m3)
```

3. What elements does it contain?
```{r}
attributes(vcov.m3)
```

4. What are the dimensions of `factors` within this object?
```{r}
dim(vcov.m3@factors$correlation)
```

# Functions
## Basics
1. Put the following math into a function $$ f(x) = 1 + 2x - 5x^2 + x^3 $$
```{r}
f <- function(x) {
  1 + 2*x - 5*x^2 + x^3
}
```

2. Set x to `1:1000/1000*6-1`
```{r}
x <- 1:1000/1000*6-1
```

3. Plot the results with 
```{r}
plot(x, f(x) , main="The answer looks like this")
```


## Scoping
1. Remove all instances of x, z, and f( ) from your environment so that you are starting fresh for this exercise.

```{r error=T}
rm(x)
rm(z)
rm(f)
```

2. What happens when we run `f()`? Why?
    - x is not defined anywhere
    
```{r eval=F} 
f <- function()
{
  return(2*x) 
}

f() 
``` 

3. What will `f()` return? Why?
    - The definition of x within the function (2) overrides the global definition of x (1)
    
```{r eval=F}
x <- 1 

f <- function()
{
  x = 2 
  return(2*x) 
}

f() 
```

4. What does the final `y` call return?
    - The defintion of y within the function does not alter the global definition of y outside of the function
    
```{r eval=F}
y <- 1

f <- function(x)
{
  y = x+2 
  return(x*y)
}

f(1)

y
```

# Building a function
## Steps 1-5
```{r}
lm.function <- function(data, cruise, x, y){
  # Load necessary packages
  require(tidyverse)
  
  # Subset the data to the cruise of interest
  dat.subset <- data %>% filter(Cruise == cruise)
  
  for(y.variable in y){ # Loop through all variables provided in y 
    # Fit a linear model 
    model <- lm(dat.subset[[y.variable]] ~ dat.subset[[x]]) 
    # Summarize the model
    sum <- summary(model)
    # Extract p-values from the summary
    pval <- sum$coefficients[,"Pr(>|t|)"]

    # Print p-values to the console
    print(pval)
  }
}
```

1. Apply the current `lm.function` to all the available geochemical variables in the Saanich data set. Which ones appear to be significantly correlated with depth?
```{r}
lm.function(data=dat, cruise=72, x="Depth", y=c("Temperature","Salinity","Density","O2","NO3","H2S"))
```

2. Copy the `lm.function` and alter it to print out the models' adjusted R-squared values instead of p-values. Be sure to run the function with inputs to make sure it works!
```{r}
lm.function.r <- function(data, cruise, x, y){
  # Load necessary packages
  require(tidyverse)
  
  # Subset the data to the cruise of interest
  dat.subset <- data %>% filter(Cruise == cruise)
  
  for(y.variable in y){ # Loop through all variables provided in y 
    # Fit a linear model 
    model <- lm(dat.subset[[y.variable]] ~ dat.subset[[x]]) 
    # Summarize the model
    sum <- summary(model)
    # Extract p-values from the summary
    r <- sum$adj.r.squared

    # Print p-values to the console
    print(r)
  }
}

lm.function.r(data=dat, cruise=72, x="Depth", y=c("O2","NO3"))
```

## Steps 6-7
1. Using our final `lm.function`, determine the linear fits for all geochemical variables for Cruise 12.
```{r}
lm.function <- function(data, cruise, x, y){
  # Load necessary packages
  require(tidyverse)
    # Remove old results file, if exists
  if(file.exists("pval_results.csv")){file.remove("pval_results.csv")}
  
  # Subset the data to the cruise of interest
  dat.subset <- data %>% filter(Cruise == cruise)
  
  for(y.variable in y){ # Loop through all variables provided in y
    # Fit a linear model 
    model <- lm(dat.subset[[y.variable]] ~ dat.subset[[x]])
    # Summarize the model
    sum <- summary(model)
    # Extract p-values from the summary
    # Reformat to a 1x2 data frame
    pval <- as.data.frame(t(sum$coefficients[,"Pr(>|t|)"]))
    # Add y variable name label
    pval$variable <- y.variable

    # Print p-values to a table
    write_csv(pval, path="pval_results.csv", append=TRUE)
  }
  # Create dynamic names for columns
  col1 <- paste(colnames(pval)[1], "p", sep=".")
  col2 <- paste(x, "p", sep=".") 
  
  # Create dynamic name fo results table
  table.name <- paste(x, "lm_pvals.csv", sep="_")
  
  # Read in p-value results and add column names
  read_csv("pval_results.csv", col_names=FALSE) %>% 
    rename(!!as.name(col1) := X1,
           !!as.name(col2) := X2,
                   variable = X3) %>%
  # Re-write the results, now with column names
  write_csv(path=table.name)
}
```

```{r}
lm.function(data=dat, cruise=72, x="Depth", y=c("Temperature","Salinity","Density","O2","NO3","H2S"))

read_csv("Depth_lm_pvals.csv")
```

2. Choose a different x variable and determine if any of the Saanich geochemical variables correlate with it.
```{r}
lm.function(data=dat, cruise=72, x="Temperature", y=c("Depth","Salinity","Density","O2","NO3","H2S"))

read_csv("Temperature_lm_pvals.csv")
```
